\section{Atomics}
\label{sect:concurrency}

Writing concurrent programs is generally considered harder than
writing sequential programs.  Similar opinions are held about
verification.  However, in VCC the leap from verifying sequential
programs to verifying fancy lock-free code is not that big.  This is
because verification in VCC is inherently based on object invariants.

Coarse-grained concurrency - where objects are protected by locks - is
really no different from sequential programming. Each lock protects a
single (typically fixed) object; when you acquire an exclusive lock,
you obtain ownership of the object protected by the lock. You are then
free to operate on tht object just as you would in sequential
programming; you can unwrap it, play with it, and eventually wrap it
up (checking that you restored the object invariant) and release the
lock, giving ownership of the protected object back to the lock.

Fine-grained concurrency - where memory is accessed using atomic
actions - is not very different from coarse-grained concurrency
(except that one is obviously restricted to operations that are
implemented atomically on the hardware platform). The
main difference is that you do not get ownership of the object you
operate on; instead, you operate on the objct by reading or writing
its volatile fields while leaving the object closed.

We begin with what is probably the most important lock free algorith:
the humble spin-lock.  The spin-lock data-structure is very simple --
it contains just a single boolean field, meant to indicate whether the
spin-lock is currently acquired. However, like most concurrent
objects, the important thing about a spinlock is forcing its users to
``play by the rules''. For example, the most important
characteristic of a spinlock is that once you acquire a spinlock, you
know that nobody else can acquire it until you release it; this would
be broken if somebody else could release the lock. But what does it
mean for ``you'' to release it? Does that mean your thread? How then
would you cope with protocols where you intentionally transfer your
right to release the spinlock to another thread?

Fortunately, we have the means at hand to represent an unforgable
right: since each object has a unique owner, we can represent a unique
right with ownership of an object. When a lock is acquired, the caller
receives ownership of an object associated with the lock (not the lock
itself, which will ultimately remain shared). To release the lock, the
caller has to give this object back. This approach prevents other
threads from competing for the lock, yet allows the thread to ``give''
his right to release the lock to another thread (or to store it in a
data structure) by simply transfering ownership.

The protected object need not serve as a simple token; being an
object, it can own data and have an invariant. These correspond to the
data ``protected'' by the lock and the notion of ``consistency'' of
this data.

\vccInput[linerange={lock-init}]{c/8.1.lockw.c}

\noindent
We use a ghost field to hold a reference to the object meant to be protected
by this lock.
If you wish to protect multiple objects with a single lock, you can make
the object referenced by \vcc{protected_obj} own them all.
The \vcc{locked} field is annotated with \vcc{volatile}.
It has the usual meaning ascribed to the modifier in C (\ie it makes the compiler
assume that the environment might write to that field, outside the knowledge
of the compiler). However, in practice C compilers also provide the 
stronger guarantee that volatile operations performed by a single
thread are guaranteed to happen in program order.

VCC treats volatile fields differently from nonvolatile fields in two
specific ways. First, a volatile field of an object can be modified
while the object is closed, as long as the update occurs inside an
explicit atomic action that preserves the object invariant. Second,
volatile fields are not in sequential domains, so a thread forgets the
values of these fields when it makes an impure function call (and, as
we will see soon, just before an atomic action).

The attribute
\vcc{_(volatile_owns)} means that we want the \vcc{\owns} set
to be treated as a volatile field (\ie we want to be able to write
it while the object is closed); without this declaration, \vcc{\owns}
sets can only be update when the object is open.

Here is the constructor for the lock:

\vccInput[linerange={init-xchg}]{c/8.1.lockw.c}

\noindent
The parameter \vcc{ob} is passed as a \Def{ghost parameter}.
The regular lock initialization function prototype does not say which
object the lock is supposed to protect, but the lock invariant requires it.
Thus, we introduce additional parameter for the purpose of verification.
A call to the initialization will look like 
\vcc{InitializeLock(&l _(ghost o))}. Note that in order to allow VCC
annotations to be erased by the preprocessor, ghost arguments are not
separated with commas.
The transfer of ownership of \vcc{ob} into the lock is exactly as in
adding an object to a container data-structure, like
\vcc{sc_add()} from \secref{dynamic-ownership}.

Now we can see how we operate on volatile fields.
We will start with the simpler of the two lock operations, the release:

\vccInput[linerange={release-out}]{c/8.1.lockw.c}

\noindent
First, let's look at the contract.
The contract above requires the lock to be wrapped, which is
hardly realistic for a lock (which of course must be shared). We will
fix this problem later, but for now, we note that what we really need
is a reliable way to make sure that the lock remains closed.


The preconditions on the protected object are very similar to the
preconditions on the \vcc{InitializeLock()}.
Note that the \vcc{Release()} does not need to mention the lock in its writes clause,
this is because the write it performs is volatile.
This is because the lock is not in the sequential domain of \vcc{\me}
(since it is a volatile field), and so the thread forgets whatever it
knew about the field when it calls a function. 

The \vcc{atomic} block is similar in spirit to the \vcc{unwrapping} block ---
it allows for modifications of listed objects and checks if their invariants
are preserved.
The difference is that the entire update happens instantaneously from the point
of view of other threads.
We needed the unwrapping operation because we wanted to mark that we temporarily
break the object invariants.
Here, there is no point in time where other threads can observe that the invariants
are broken.
Invariants hold before the beginning of the atomic block (by our principal reasoning
rule, \secref{wrap-unwrap}), and we check the invariant at the end of the atomic block.

For the use of an atomic action construct to be sound, VCC requires
that the actions on volatile concrete data of closed objects performed
by the action must appear atomic to other threads.  VCC issues a
warning if there is more than one volatile physical memory operation
inside of an atomic block (since this is unlikely to be atomic on most
architectures), but is otherwise silent, assuming that a single
volatile memory access is implemented in an atomic way by the host
compiler (even though this is not true for all data types on all
platforms). Ultimately,  it is up to the user to
guarantee that the access performed is indeed atomic.

Here, the physically atomic operation is writing to the \vcc{l->locked}; Other
possibilities include reading from a volatile field, or a performing a
primitive operation supported by the hardware, like interlocked
compare-and-exchange, on such a field.  The block may also include any
number of accesses to mutable variables, reads of thread-local
variables, reads of nonvolatile fields of objects named in the atomic,
reads of volatile ghost variables (in any object), and writes of
volatile ghost variables of objects named in the atomic. The
additional operations on concrete fields are allowed because these
accesses are not racing with other threads, and the ghost operations
are allowed because we can pretend that ghost updates happen
immediately (without scheduler boundary).

It is not hard to see that this atomic operation preserves the
invariant of the lock.
However, VCC must also make the usual checks on the sequential data.
To transfer ownership of \vcc{l->protected_obj} to the lock, we also need
write permission to the object being transferred, and
we need to know it is closed.
For example, had we forgotten to mention \vcc{l->protected_obj}
in the writes clause,  VCC would have complained:

\vccInput[linerange={out-999}]{c/8.2.lockw_wrong.c}

\noindent
And had we forgotten to perform the ownership transfer inside of
\vcc{Release()}, VCC would have complained
about the invariant of the lock:

\vccInput[linerange={out-999}]{c/8.3.lockw_wrong2.c}

Now we'll look at \vcc{Acquire()}. 
The specification is not very surprising. It requires the lock to be
wrapped (again, unrealistic for a shared lock, but we will fix this later).
It ensures that the calling thread will own the protected object,
and moreover, that the thread didn't directly own it before.
This is much like the postcondition on \vcc{sc_set()} function
from \secref{dynamic-ownership}.

\vccInput[linerange={acquire-release}]{c/8.1.lockw.c}

\noindent
The \vcc{InterlockedCompareAndExchange()} function is a compiler instrinsic,
which on the x86/x64 hardware translates to the \vcc{cmpxchg} assembly instruction.
It takes a memory location and two values.
If the memory location contains the first value, then it is replaced with the second.
It returns the old value.
The entire operation is performed atomically (and is also a write barrier).

VCC doesn't have all the primitives of all the C compilers predefined.
You define such intrinsics for VCC by providing a body. 
It is seen only by VCC (it is enclosed in
\vcc{_(atomic_inline ...)}) so that the normal compiler doesn't get confused
about it.

\vccInput[linerange={xchg-acquire}]{c/8.1.lockw.c}

\noindent
It is up to you to make sure that any such functions you provide
indeed match the semantics provided by your compiler and platform. 
Such definitions may in future be provided in header files for certain compiler/platform
combinations.


\subsection{Using claims}
\label{sect:using-claims}

The contracts of functions operating on the lock require that the lock
is wrapped.
This is because one can only perform atomic operations on objects
that are closed. 
If an object is open, then the owning thread is in full control of it.
However, wrapped means not only closed, but also owned by the current thread,
which defeats the purpose of the lock --- it should be possible
for multiple threads to compete for the lock.
Let's then say, there is a thread which owns the lock.
Assume some other thread \vcc|t| got to know that the lock is closed.
How would \vcc|t| know that the owning thread won't unwrap (or worse yet, deallocate) the lock, just
before \vcc|t| tries an atomic operation on the lock?
The owning thread thus needs to somehow promise \vcc|t|
that lock will stay closed.
In VCC such a promise takes the form of a \Def{claim}.
Later we'll see that claims are more powerful, but for
now consider the following to be the definition of a claim:

\begin{VCC}
_(ghost 
typedef struct {
  \ptrset claimed;
  _(invariant \forall \object o; o \in claimed ==> o->\closed)
} \claim_struct, *\claim;
)
\end{VCC}

\noindent
Thus, a claim is an object, with an invariant stating that a number of other objects
(we call them \Def{claimed objects}) are closed.
As this is stated in the invariant of the claim, it only needs to be true
as long as the claim itself stays closed.

Recall that what can be written in invariants is subject to the admissibility
condition, which we have seen partially explained in \secref{admissibility0}.
There we said that an invariant should talk only about things the object owns.
But here the claim doesn't own the claimed objects,
so how should the claim know the object will stay closed?
In general, an admissible invariant can depend on other objects invariants always being
preserved (we'll see the precise rule in \secref{inv2}).
So VCC adds an implicit invariant to all types
marked with \vcc{_(claimable)} attribute.
This invariant states that the object cannot be unwrapped when
there are closed claims on it.
More precisely, each claimable object keeps track of the count of outstanding
claims.
The number of outstanding claims on an object is stored in
\vcc{\claim_count} field.

Now, getting back to our lock example, the trick is that there can be
multiple claims claiming the lock (note that this is orthogonal to
the fact that a single claim can claim multiple objects).
The thread that owns the lock will need to keep track of who's using
the lock.
The owner won't be able to destroy the lock (which requires unwrapping it),
before it makes sure there is no one using the lock.
Thus, we need to add \vcc{_(claimable)} attribute to our lock
definition, and change the contract on the functions operating
on the lock. As the changes are very similar we'll only
show \vcc{Release()}.

\vccInput[linerange={release-struct_data}]{c/8.4.lock_claimsobj.c}

\noindent
We pass a ghost parameter holding a claim.
The claim should be wrapped.
The function \vcc{\claims_obj(c, l)} is defined to be
\vcc{l \in c->claimed}, \ie that the claim claims the lock.
We also need to know that the claim is not the protected object,
otherwise we couldn't ensure that the claim is wrapped after the call.
This is the kind of weird corner case that VCC is very good catching
(even if it's bogus in this context).
Other than the contract, the only change is that we list the claim
as parameter to the atomic block.
Listing a normal object as parameter to the atomic makes VCC know you're
going to modify the object.
For claims, it is just a hint, that it should use this claim when trying
to prove that the object is closed.

Additionally, the \vcc{InitializeLock()} needs to ensure \vcc{l->\claim_count} \vcc{== 0}
(\ie no claims on freshly initialized lock).
VCC even provides a syntax to say something is wrapped and has no claims: \vcc{\wrapped0(l)}.

\subsection{Creating claims}
\label{sect:creating-claims}

When creating (or destroying) a claim one needs to list the claimed objects.
Let's have a look at an example.

\vccInput[linerange={create_claim-out}]{c/8.4.lock_claimsobj.c}

This function tests that we can actually create a lock, create a claim on it,
use the lock, and then destroy it.
The \vcc{InitializeLock()} leaves the lock wrapped and writable by the current thread.
This allows for the creation of an appropriate claim, which is then passed to \vcc{Acquire()} and \vcc{Release()}.
Finally, we destroy the claim, which allows for unwrapping of the lock, and subsequently deallocating
it when the function activation record is popped off the stack.

The \vcc{\make_claim(...)} function takes the set of objects to be claimed
and a property (an invariant of the claim, we'll get to that in the next section).
Let us give desugaring of \vcc{\make_claim(...)} for a single object
in terms of the \vcc{\claim_struct} defined in the previous section.

\begin{VCC}
// c = \make_claim({o}, \true) expands to
o->\claim_count += 1;
c = malloc(sizeof(\claim_struct));
c->claimed = {o};
_(wrap c);

// \destroy_claim(c, {o}) expands to
assert(o \in c->claimed);
o->\claim_count -= 1;
_(unwrap c);
free(c);
\end{VCC}


Because creating or destroying a claim on \vcc{c} assigns to
\vcc{c->\claim_count}, it requires write access to that memory location.
One way to obtain such access is getting sequential write access to \vcc{c} itself:
in our example the lock is created on the stack and thus sequentially writable.
We can thus create a claim and immediately use it.
A more realistic claim management scenario is described in \secref{dynamic-claims}.
%when a thread creates an object, constructs
%a number of claims on it, and stores the claims in some shared, possibly global, data-structures
%(\eg a work-queue) where other threads can access them.

The \vcc{\true} in \vcc{\make_claim(...)} is the claimed property (an invariant
of the claim), which will be explained in the next section.

\begin{note}
The destruction can possibly leak claim counts, \ie one could say:
\begin{VCC}
\destroy_claim(c, {});
\end{VCC}
\noindent
and it would verify just fine.
This avoids the need to have write access to \vcc{p}, but on the other hand prevents
\vcc{p} from unwrapping forever (which might be actually fine if \vcc{p} is a ghost object).
%It seems clear why the claimed objects need to be listed when creating a claim, but
%why do we need them for destruction?
\end{note}

\subsection{Two-state invariants}
\label{sect:inv2}

Sometimes it is not only important what are the valid states of objects,
but also what are the allowed \emph{changes} to objects.
For example, let's take a counter keeping track of certain operations
since the beginning of the program.

\vccInput[linerange={counter-reading}]{c/8.5.counter.c}

\noindent
Its first invariant is a plain single-state invariant -- for some reason
we decided to exclude zero as the valid count.
The second invariant says that for any atomic update of (closed)
counter, \vcc{v} can either stay unchanged or increment by exactly one.
The syntax \vcc{\old(v)} is used to refer to value of \vcc{v} before
an atomic update, and plain \vcc{v} is used for the value of \vcc{v}
after the update.
(Note that the argument to \vcc{\old(...)} can be an arbitrary expression.)
That is, when checking that an atomic update preserves the invariant
of a counter, we will take the state of the program right
before the update, the state right after the update, and check
that the invariant holds for that pair of states.

\begin{note}
In fact, it would be easy to prevent any changes to some field \vcc{f}, by
saying \vcc{_(invariant \old(f) == f)}.
This is roughly what happens under the hood when a field is
declared without the \vcc{volatile} modifier.
\end{note}

As we can see the single- and two-state invariants are both defined
using the \vcc{_(invariant ...)} syntax.
The single-state invariants are just two-state invariants, which do not use
\vcc{\old(...)}.
However, we often need an interpretation of an object invariant in a single state \vcc{S}.
For that we use the \Def{stuttering transition} from \vcc{S} to \vcc{S} itself.
VCC enforces that all invariants are \Def{reflexive} that is if they hold
over a transition \vcc{S0, S1}, then they should hold in just \vcc{S1}
(\ie over \vcc{S1, S1}).
In practice,
this means that \vcc{\old(...)} should be only used to describe
how objects change, and not what are their proper values.
In particular,
all invariants which do not use \vcc{\old(...)} are reflexive, and so
are all invariants of the form \vcc{\old(E) == (E) || (P)}, for any expression \vcc{E} and condition \vcc{P}.
On the other hand, the invariants \vcc{\old(f) < 7} and \vcc{x == \old(x) + 1} are not reflexive.

Let's now discuss where can you actually rely on invariants being preserved.

\begin{VCC}
void foo(struct Counter *n)
  _(requires \wrapped(n))
{
  int x, y;
  atomic(n) { x = n->v; }
  atomic(n) { y = n->v; }
}
\end{VCC}

\noindent
The question is what do we know about \vcc{x} and \vcc{y}
at the end of \vcc{foo()}.
If we knew that nobody is updating \vcc{n->v} while \vcc{foo()}
is running we would know \vcc{x == y}.
This would be the case if \vcc{n} was unwrapped, but it is wrapped.
In our case, because \vcc{n} is closed, other threads can update it,
while \vcc{foo()} is running, but they will need to
adhere to \vcc{n}'s invariant.
So we might guess that at end of \vcc{foo()} we know
\vcc{y == x || y == x + 1}.
But this is incorrect: \vcc{n->v} might get incremented
by more than one, in several steps.
The correct answer is thus \vcc{x <= y}.
Unfortunately, in general, such properties are very difficult to deduce
automatically, which is why we use plain object invariants and admissibility
check to express such properties in VCC.

\begin{note}
An invariant is \Def{transitive} if it holds over states \vcc{S0, S2},
provided that it holds over \vcc{S0, S1} and \vcc{S1, S2}.
Transitive invariants could be assumed over arbitrary
pairs of states, provided that the object stays closed
in between them. 
VCC does not require invariants to be transitive, though.

Some invariants are naturally transitive (\eg we could say
\vcc{_(invariant \old(x) <= x)} in \vcc{struct Counter},
and it would be almost as good our current invariant).
Some other invariants, especially the more complicated ones,
are more difficult to make transitive.
For example, an invariant on a reader-writer lock might say
\begin{VCC}
_(invariant writer_waiting ==> old(readers) >= readers)
\end{VCC}
\noindent
To make it transitive one needs to introduce version numbers.
Some invariants describing hardware (\eg a step of physical CPU)
are impossible to make transitive.
\end{note}

Consider the following structure definition:

\vccInput[linerange={reading-endreading}]{c/8.5.counter.c}

\noindent 
It is meant to represent a reading from a counter.
Let's consider its admissibility.
It has a pointer to the counter, and a owns a claim, which
claims the counter.
So far, so good.
It also states that the current value of the counter is no less than \vcc{r}.
Clearly, the \vcc{Reading} doesn't own the counter, so our previous rule
from \secref{admissibility0}, which states
that you can mention in your invariant everything that you own, doesn't apply.
It would be tempting to extend that rule to say ``everything that you own
or have a claim on'', but VCC actually uses a more general rule.
In a nutshell, the rule says that every invariant should be preserved
under changes to other objects, provided that these other objects change
according to their invariants.
When we look at our \vcc{struct Reading}, its invariant cannot be broken when
its counter increments, which is the only change allowed by counters invariant.
On the other hand, an invariant like \vcc{r == n->v} or \vcc{r >= n->v}
could be broken by such a change.
But let us proceed with somewhat more precise definitions.

First, assume that every object invariant holds when the object is not closed.
This might sound counter-intuitive, but remember that closedness is controlled
by a field.
When that field is set to false, we want to \emph{effectively} disable the invariant,
which is the same as just forcing it to be true in that case.
Alternatively, you might try to think of all objects as being closed for a while.

An atomic action, which updates state \vcc{S0} into \vcc{S1}, is \Def{legal} if and only if the invariants of
objects that have changed between \vcc{S0} and \vcc{S1} hold over \vcc{S0, S1}.
In other words, a legal action preservers invariants of updated objects.
This should not come as a surprise: this is exactly what VCC checks
for in atomic blocks.

An invariant is \Def{stable} if and only if it cannot be broken by legal updates.
More precisely, to prove that an invariant of \vcc{p} is stable,
VCC needs to ``simulate'' an arbitrary legal update:
\begin{itemize}
\item Take two arbitrary states \vcc{S0} and \vcc{S1}.
\item Assume that all invariants (including \vcc{p}'s) hold over \vcc{S0, S0}.
\item Assume that for all objects, some fields of which are not the same in \vcc{S0} and \vcc{S1},
their invariants hold over \vcc{S0, S1}.
\item Assume that all fields of \vcc{p} are the same in \vcc{S0} and \vcc{S1}.
\item Check that invariant of \vcc{p} holds over \vcc{S0, S1}.
\end{itemize}
The first assumption comes from the fact that all invariants are reflexive.
The second assumption is legality.
The third assumption follows from the second (if \vcc{p} did change, its invariant would
automatically hold).

An invariant is \Def{admissible} if and only if it is stable and reflexive.

Let's see how our previous notion of admissibility relates to this one.
If \vcc{p} owns \vcc{q}, then \vcc{q \in p->\owns}.
By the third admissibility assumption, after the simulated action \vcc{p} still owns \vcc{q}.
By the rules of ownership (\secref{wrap-unwrap}), only threads can own
open objects, so we know that \vcc{q} is closed in both \vcc{S0}
and \vcc{S1}.
Therefore non-volatile fields of \vcc{q} do not change between \vcc{S0} and \vcc{S1},
and thus the invariant of \vcc{p} can freely talk about their values:
whatever property of them was true in \vcc{S0}, will also be true in \vcc{S1}.
Additionally, if \vcc{q} owned \vcc{r} before the atomic action, and the \vcc{q->\owns} is non-volatile,
it will keep owning \vcc{r}, and thus non-volatile fields of \vcc{r}
will stay unchanged.
Thus our previous notion of admissibility is a special case of this one.

Getting back to our \vcc{foo()} example, to deduce that \vcc{x <= y}, after
the first read we could create a ghost \vcc{Reading} object, and
use its invariant in the second action.
While we need to say that \vcc{x <= y} is what's required,
using a full-fledged object might seem like an overkill.
Luckily, definitions of claims themselves can specify additional invariants.

\begin{note}
The admissibility condition above is semantic: it will be checked by the theorem
prover. 
This allows construction of the derived concepts like claims and ownership,
and also escaping their limitations if needed.
It is therefore the most central concept of VCC verification methodology,
even if it doesn't look like much at the first sight.
\end{note}

\subsection{Guaranteed properties in claims}
\label{sect:claim-props}

When constructing a claim, you can specify additional invariants to put on
the imaginary definition of the claim structure.
Let's have a look at annotated version of our previous \vcc{foo()} function.

\vccInput[linerange={readtwice-endreadtwice}]{c/8.5.counter.c}

\noindent
Let's give a high-level description of what's going on.
Just after reading \vcc{n->v} we create a claim \vcc{r}, which guarantees
that in every state, where \vcc{r} is closed,
the current value of \vcc{n->v} is no less than the value of \vcc{x}
at the time when \vcc{r} was created.
Then, after reading \vcc{n->v} for the second time, we tell VCC to
make use of \vcc{r}'s guaranteed property, by asserting that it is ``active''.
This makes VCC know \vcc{x <= n->v} in the current state, where also
\vcc{y == n->v}.
From these two facts VCC can conclude that \vcc{x <= y}.

The general syntax for constructing a claim is:

\begin{VCC}
_(ghost c = \make_claim(S, P))
\end{VCC}

\noindent
We already explained, that this requires that \vcc{s->\claim_count} is writable for \vcc{s \in S}.
As for the property \vcc{P}, we pretend it forms the invariant of the claim.
Because we're just constructing the claim, just like during regular object initialization,
the invariant has to hold initially (\ie at the moment when the claim is created,
that is wrapped).
Moreover, the invariant has to be admissible, under the condition
that all objects in \vcc{S} stay closed as long as the claim itself
stays closed.
The claimed property cannot use \vcc{\old(...)}, and therefore it's automatically
reflexive, thus it only needs to be stable to guarantee admissibility.

But what about locals?
Normally, object invariants are not allowed to reference locals.
The idea is that when the claim is constructed, all the locals that the
claim references are copied into imaginary fields of the claim.
The fields of the claim never change, once it is created.
Therefore an assignment \vcc{x = UINT_MAX;} in between the atomic
blocks would not invalidate the claim --- the claim would still
refer to the old value of \vcc{x}.
Of course, it would invalidate the final \vcc{x <= y} assert.

\begin{note}
For any expression \vcc{E} you can use \vcc{\at(\now(), E)} in \vcc{P}
in order to have the value of \vcc{E} be evaluated in the state
when the claim is created, and stored in the field of the claim.
\end{note}

This copying business doesn't affect initial checking of the \vcc{P},
\vcc{P} should just hold at the point when the claim is created.
It does however affect the admissibility check for \vcc{P}:
\begin{itemize}
\item Consider an arbitrary legal action, from \vcc{S0} to \vcc{S1}.
\item Assume that all invariants hold over \vcc{S0, S0}, including assuming \vcc{P} in \vcc{S0}.
\item Assume that fields of \vcc{c} didn't change between \vcc{S0} and \vcc{S1}
(in particular locals referenced by the claim are the same as at the moment of its creation).
\item Assume all objects in \vcc{S} are closed in both \vcc{S0} and \vcc{S1}.
\item Assume that for all objects, fields of which are not the same in \vcc{S0} and \vcc{S1},
their invariants hold over \vcc{S0, S1}.
\item Check that \vcc{P} holds in \vcc{S1}.
\end{itemize}

To prove \vcc{\active_claim(c)} one needs to prove \vcc{c->\closed} and that
the current state is a \Def{full-stop} state, \ie state where all invariants
are guaranteed to hold.
Any execution state outside of an atomic block is full-stop.
The state right at the beginning of an atomic block is also full-stop.
The states in the middle of it (\ie after some state updates) might not be.

\begin{note}
Such middle-of-the-atomic states are not observable by other threads, and therefore
the fact that the invariants don't hold there does not create soundness problems.
\end{note}

The fact that \vcc{P} follows from \vcc{c}'s invariant after the construction
is expressed using \vcc{\claims(c, P)}.
It is roughly equivalent to saying:
\begin{VCC}
\forall \state s {\at(s, \active_claim(c))};
  \at(s, \active_claim(c)) ==> \at(s, P)
\end{VCC}
Thus, after asserting \vcc{\active_claim(c)} in some state \vcc{s},
\vcc{\at(s, P)} will be assumed, which means VCC will
assume \vcc{P}, where all heap references are replaced by their values in
\vcc{s}, and all locals are replaced by the values at the point
when the claim was created.

\itodo{I think we need more examples about that at() business,
claim admissibility checks and so forth}

\subsection{Dynamic claim management}
\label{sect:dynamic-claims}

So far we have only considered the case of creating claims to wrapped objects.
In real systems some resources are managed dynamically:
threads ask for ``handles'' to resources, operate on them,
and give the handles back.
These handles are usually purely virtual --- asking for a handle amounts to incrementing
some counter.
Only after all handles are given back the resource can be disposed.
This is pretty much how claims work in VCC, and indeed they were modeled after this
real-world scenario. 
Below we have an example of prototypical reference counter.

\vccInput[linerange={refcnt-init}]{c/8.6.rundown.c}

\noindent
Thus, a \vcc{struct RefCnt} owns a resource, and makes sure that the number of outstanding
claims on the resource matches the physical counter stored in it.
\vcc{\claimable(p)} means that the type of object pointed to by \vcc{p} was marked
with \vcc{_(claimable)}.
The lowest bit is used to disable giving out of new references
(this is expressed in the last invariant).

\vccInput[linerange={init-incr}]{c/8.6.rundown.c}

\noindent
Initialization shouldn't be very surprising:
\vcc{\wrapped0(o)} means \vcc{\wrapped(o) && o->\claim_count == 0},
and thus on initialization we require a resource without any outstanding
claims.

\vccInput[linerange={incr-decr}]{c/8.6.rundown.c}

\noindent
First, let's have a look at the function contract.
The syntax \vcc{_(always c, P)} is equivalent to:
\begin{VCC}
  _(requires \wrapped(c) && \claims(c, P))
  _(ensures \wrapped(c))
\end{VCC}
Thus, instead of requiring \vcc{\claims_obj(c, r)}, we require that the claim
guarantees \vcc{r->\closed}.
One way of doing this is claiming \vcc{r}, but another is claiming the owner
of \vcc{r}, as we will see shortly.

As for the body, we assume our reference counter will never overflow.
This clearly depends on the running time of the system and usage patterns,
but in general it would be difficult to specify this, and thus we just
hand-wave it.

The new thing about the body is that we make a claim on the resource,
even though it's not wrapped.
There are two ways of obtaining write access to \vcc{p->\claim_count}:
either having \vcc{p} writable sequentially and wrapped,
or in case \vcc{p->\owner} is a non-thread object, checking
invariant of \vcc{p->\owner}.
Thus, inside an atomic update on \vcc{p->\owner} (which will check the invariant of \vcc{p->\owner}) one can create
claims on \vcc{p}.
The same rule applies to claim destruction:

\vccInput[linerange={decr-use}]{c/8.6.rundown.c}

\noindent
A little tricky thing here, is that we need to make use of the \vcc{handle} claim
right after reading \vcc{r->cnt}. 
Because this claim is valid, we know that the claim count on the resource
is positive and therefore (by reference counter invariant) \vcc{v >= 2}.
Without using the \vcc{handle} claim to deduce it we would get a complaint
about overflow in \vcc{v - 2} in the second atomic block.

Finally, let's have a look at a possible use scenario of our reference counter.

\vccInput[linerange={use-enduse}]{c/8.6.rundown.c}

\noindent
The \vcc{struct B} contains a \vcc{struct A} governed by a reference counter.
It owns the reference counter, but not \vcc{struct A} (which is owned by the reference
counter).
A claim guaranteeing that \vcc{struct B} is closed also guarantees
that its counter is closed, so we can pass it to \vcc{try_incr()},
which gives us a handle on \vcc{struct A}.

Of course a question arises where one does get a claim on \vcc{struct B} from?
In real systems the top-level claims come either from global objects that are
always closed, or from data passed when the thread is created.

